---
title: 'Data Analysis #1'
author: "Dustin Leatherman"
date: "February 9, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(openintro)
library(tidyr)
library(dplyr)
library(broom)

library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)

# done moreso for graphing purposes
bdimsWithGender <- 
  bdims %>% 
  mutate(gender = ifelse(sex == 1, "male", "female")) 
```

# Body Girth and Physical Activity

## Introduction

Body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender, were measured for 507 physically active individuals - 247 men and 260 women. This dataset was initially published in the article, "Exploring Relationships in Body Dimensions" in the *Journal of Statistics Education*. 

Many measurements were taken but the researchers are particularly interested in predicting an individual's weight in kilograms from just their waist girth in centimeters. Even so, it is beneficial to compare weight with all other measurements to see why the researchers made their decision.

```{r summaryPlots1, echo=FALSE, fig.width=10, fig.align='center', fig.height=8}

# Show all girth-related scatterplots with relation to weight since that is the response of interest. Color-coded by gender
# Scales are set to free since measurements on body parts vary greatly
bdimsWithGender %>%
  select(
    Weight = wgt, 
    Gender = gender, 
    Chest = che.gi,
    Shoulder = sho.gi, 
    Bicep = bic.gi, 
    Forcep = for.gi,
    Wrist = wri.gi,
    Waist = wai.gi, 
    Navel = nav.gi, 
    Hip = hip.gi, 
    Thigh = thi.gi,
    Knee = kne.gi, 
    Calf = cal.gi, 
    Ankle = ank.gi
    ) %>% 
  # remove wgt and gender from gathered data. We don't want to include these as explanatory variables in our plots
  gather(c(-Weight,-Gender), key = "variable", value = "value") %>% 
  ggplot(aes(x = value, y = Weight, color = Gender)) + 
    geom_point(alpha = 0.6) + 
    facet_wrap(~ variable , ncol = 3, scales = "free") +
      labs(color = "Gender", subtitle = "All Girth measurements compared against weight, broken down by Gender") +
      ylab("Weight (kg)") +
      xlab("Girth (cm)") +
      ggtitle("Comparing Weight and Girth") +
        theme(plot.title = element_text(hjust = 0.5))
```


```{r summaryPlots2, fig.width=10, fig.align='center', fig.height=8}
# Show all diameter-related scatterplots with relation to weight. Color-coded by gender.
bdimsWithGender %>%
  select(
    Weight = wgt, 
    Gender = gender, 
    Biacromial = bia.di, 
    Biiliac = bii.di, 
    Bitrochanteric = bit.di, 
    `Chest Depth` = che.de, 
    Chest = che.di, 
    Elbow = elb.di, 
    Wrist = wri.di, 
    Knee = kne.di, 
    Ankle = ank.di) %>% 
  # remove wgt and gender from gathered data. We don't want to include these as explanatory variables in our plots
  gather(c(-Weight,-Gender), key = "variable", value = "value") %>% 
  ggplot(aes(x = value, y = Weight, color = Gender)) + 
    geom_point(alpha = 0.6) + 
    facet_wrap(~ variable , ncol = 3, scales = "free") +
      labs(color = "Gender", subtitle = "All Diameter measurements compared against weight, broken down by Gender.") +
      ylab("Weight (kg)") +
      xlab("Diameter (cm)") +
      ggtitle("Comparing Weight and Diameter") +
        theme(plot.title = element_text(hjust = 0.5))
```


```{r summaryPlots3, fig.width=10, fig.align='center'}
# Show the rest of the variables of interest. In this case, it is only age and height
plot.height <- bdimsWithGender %>%
  select(Weight = wgt, Gender = gender, Height = hgt) %>% 
  ggplot(aes(x = Height, y = Weight, color = Gender)) + 
    geom_point(alpha = 0.6) + 
      ylab("Weight (kg)") +
      xlab("Height (cm)") +
      ggtitle("Comparing Weight and Height") +
        theme(plot.title = element_text(hjust = 0.5))

plot.age <- bdimsWithGender %>%
  select(Weight = wgt, Gender = gender, Age = age) %>% 
  ggplot(aes(x = Age, y = Weight, color = Gender)) + 
    geom_point(alpha = 0.6, show.legend = FALSE) + 
      ylab("Weight (kg)") +
      xlab("Age (years)") +
      ggtitle("Comparing Weight and Age") +
        theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plot.age, plot.height,
               widths = c(1,1),
               ncol = 2)
```

All girth and diameter explanatory variables have a positive correlation with weight. It appears that girth of body parts in general may have a stronger correlation to weight than diameter. This can be determined from the more focused trendline in the girth plots than the diameter plots. In all cases, there appears to be a clear distinction between males and females in terms of weight. Males appear to be heavier than females for all measurements barring age. There are two visible patterns within the graphs. The first being those which females and males have little overlapping variables. Examples of this include Forcep Girth vs Weight, Chest Girth vs Weight, and Wrist Girth vs Weight. The second pattern is that for a male and female of the same measurement, the male is generally heavier. This may indicate that this measurement is not as impactful as one where there are fewer overlapping values. Examples of this include Thigh Girth vs Weight, Calf Girth vs Weight, and Bicep Diameter vs Weight. Waist Girth vs Weight has a fairly clear and focused trendline. There appears to be a decent amount of intersection between males and females in the middle so it is understandable why this is the main variable of interest.

The simple linear regression model they have used is as follows:

  $\mu$ {Weight|Waist girth} = $\beta_0$ + $\beta_1$Waist girth
    
This linear model does not take gender into account. As seen in the scatterplots, gender appears to be associated with weight so the researchers are interested in whether they should be using different lines for men and women.

```{r waistGirthScatter, fig.width=10, fig.align='center'}
bdimsWithGender %>%
  select("wgt", "wai.gi", "gender") %>% 
  ggplot(aes(x = wai.gi, y = wgt, color = gender)) + 
    geom_point(alpha = 0.6) +
      labs(color = "Gender", subtitle = "Participants marked by their respective genders") +
      ylab("Weight (kg)") +
      xlab("Waist Girth (cm)") +
      ggtitle("Comparing Weight and Waist Girth") +
        theme(plot.title = element_text(hjust = 0.5))

```

These are our questions of interest:

* After taking into account Waist Girth, is the deviation in weight between men and women variable or fixed?

* Based on what is known, what is the most appropriate model for estimating average weight?

# Methods

For convenience, the single parameter model is as follows:

$\mu$ {Weight|Waist girth} = $\beta_0$ + $\beta_1$Waist girth

The most common method to create a separate lines model is to add an interaction term. Since the researchers are interested in the deviation between men and women, gender should be added as an indicator variable as well as an interaction term with waist girth. 

The separate lines model is as follows:

$\mu$ {Weight|Waist girth, MALE} = $\beta_0$ + $\beta_1$Waist girth + $\beta_2$MALE + $\beta_3$(MALE $\times$ Waist girth)


```{r setupmodels, fig.width=10, warning=FALSE, message=FALSE}

# canned waist girth values per gender
new_data <- expand.grid(wai.gi = c(60,80,100), gender = unique(bdimsWithGender$gender))

# I tried for longer than I care to admit to make models work within the tidyverse framework
# I was able to get it to run just fine in my console but would run into strange issues when 
# generating the report here. It's a shame because I really like Tidyverse and the principles
# behind it. Alas, there must be some error (user or otherwise) that makes it not play nicely with RMarkdown.
model.initial <- lm(wgt ~ wai.gi, data = bdimsWithGender)
model.initial.conf <- confint_tidy(model.initial)
model.initial.conf.pred <- predict(model.initial, newdata = new_data, interval = "confidence")

model.separate_lines <- lm(wgt ~ wai.gi * gender, data = bdimsWithGender)
model.separate_lines.conf <- confint_tidy(model.separate_lines)
model.separate_lines.conf.pred <- predict(model.separate_lines, newdata = new_data, interval = "confidence")
```


```{r displayModel21, fig.width=10, warning=FALSE, message=FALSE, fig.pos="!h"}

kable(
  new_data %>% 
    bind_cols(
      model.initial.conf.pred %>% 
        tidy, 
      model.separate_lines.conf.pred %>% 
        tidy
      ) %>% 
    arrange(wai.gi),
  align = c('l'),
  digits = 4,
  col.names = c("Waist Girth (cm)","Gender","Fit", "Lower", "Upper", "Fit", "Lower", "Upper"),
  caption = "Fit + Confidence Interval for specified values of Waist Girth"
) %>% 
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  add_header_above(c(" " = 2, "Single Parameter (kg)" = 3, "Separate Lines (kg)" = 3)) %>%
  row_spec(0, bold = T)

```


The confidence intervals for the mean weight after taking into account Waist Girth is the same between genders for the Single Parameter model. This is expected since gender is not a parameter in the Single Parameter model. The confidence intervals for the Separate Lines model varies between gender after taking into account Waist Girth. These confidence intervals appear to be wider than the Single Parameter model. From a non-technical perspective, this means that by including Gender and its interaction with Waist Girth, there is more variability in expected results for a given waist girth and gender. 

The range appears to be wider at Waist Girth values of 60 and 100 than 80 cm which indicates that there is more variability in the tails of the Separate line model . The Waist Girth tails (60 and 100cm) of the Single Parameter model are wider than the center value (80cm) but not as wide as the Separate Lines model.

# Results

## Linear Terms

There is moderate evidence in the Separate Lines model that Waist Girth reduces weight for males compared to females.

```{r modelOutput, fig.width=10, warning=FALSE, message=FALSE}
kable(
  tidy(model.initial),
  align = c('l'),
  digits = 4,
  col.names = c("Explanatory Variable","Fit (kg)","Standard Error (kg)", "T-Statistic", "p-value"),
  caption = "Single Parameter Model"
) %>%
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  column_spec(1:1, bold = T, border_right = T) %>%
  row_spec(0, bold = T)
```


```{r modelOutput2, fig.width=10, warning=FALSE, message=FALSE}
kable(
  tidy(model.separate_lines),
  align = c('l'),
  digits = 4,
  col.names = c("Explanatory Variable","Fit (kg)","Standard Error (kg)", "T-Statistic", "p-value"),
  caption = "Separate Lines Model"
) %>%
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  column_spec(1:1, bold = T, border_right = T) %>%
  row_spec(0, bold = T)
```


```{r modelCompare, warning=FALSE, message=FALSE}
kable (
  anova(model.initial, model.separate_lines) %>% tidy,
  align = c('l'),
  digits = 4,
  #col.names = c("Explanatory Variable","Fit (kg)","Standard Error (kg)", "T-Statistic", "p-value"),
  caption = "Sum of Squares F-Test to compare Single Parameter and Separate Lines Model"
) %>%
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  row_spec(0, bold = T)
```

There is convincing evidence that the Single Parameter model does not have a better fit than the Separate Lines model (p-value = 0.0001). 

```{r finalModelPlot, fig.width=10, warning=FALSE, message=FALSE}
model.separate_lines %>%
  ggplot(aes(x = wai.gi, y = wgt, color = gender)) + 
    geom_point(alpha = 0.6) +
    geom_line(aes(wai.gi, .fitted), size = 1) +
      labs(color = "Gender", subtitle = "Participants marked by their respective genders") +
      ylab("Weight (kg)") +
      xlab("Waist Girth (cm)") +
      ggtitle("Comparing Weight and Waist Girth for Men and Women") +
        theme(plot.title = element_text(hjust = 0.5))
```

The Separate Lines model is the better for estimating weight with Waist Girth than the Single Parameter model initially proposed by the researchers. More analysis is needed as there appears to be some outliers and influencers that affect the slopes. 

## Quadratic Terms

The male weight appears to curve downwards for the high Waist Girth values. The Separate Lines model indicates that this is the case since the interaction term produced a negative value. It is worth seeing whether or not a quadratic term for Waist Girth provides a better fit. 


```{r quadraticModels, fig.width=10, warning=FALSE, message=FALSE}
model.separate_lines.poly <- lm(wgt ~ poly(wai.gi, 2) * gender, data = bdimsWithGender) 

# Display Fit and Summary Information
kable(
  model.separate_lines.poly %>% tidy,
  align = c('l'),
  digits = 4,
  col.names = c("Explanatory Variable","Fit (kg)","Standard Error (kg)", "T-Statistic", "p-value"),
  caption = "Separate Lines with Quadratic Waist Girth"
) %>%
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  column_spec(1:1, bold = T, border_right = T) %>%
  row_spec(0, bold = T)

# Display SS F-Test Table
kable (
  anova(model.separate_lines, model.separate_lines.poly) %>% tidy,
  align = c('l'),
  digits = 4,
  caption = "Sum of Squares F-Test to compare Separate Lines and Separate Lines Quadratic Model"
) %>%
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
  row_spec(0, bold = T)

# For some reason, this renders fine in my console but not through here
# This function is preferred since it reuses the model but I couldn't figure it out.
# The plot below this works as expected.

# ggplot(model.separate_lines.poly, aes(x = wai.gi, y = wgt, color = gender)) +
#   geom_point(alpha = 0.6) +
#   geom_line(aes(y = .fitted)) +
#     labs(color = "Gender", subtitle = "Participants marked by their respective genders") +
#     ylab("Weight (kg)") +
#     xlab("Waist Girth (cm)") +
#     ggtitle("Comparing Weight and Waist Girth for Men and Women") +
#       theme(plot.title = element_text(hjust = 0.5))

# Graph the polynomial model against the scatterplot
ggplot(bdimsWithGender) +
  geom_point(aes(x = wai.gi, y = wgt, color = gender), alpha = 0.6) +
    # gender isnt included but comparing the graphs, it doesnt affect the line that is drawn so it should be okay.
    stat_smooth(aes(x = wai.gi, y = wgt, color = gender), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
    labs(color = "Gender", subtitle = "Participants marked by their respective genders") +
    ylab("Weight (kg)") +
    xlab("Waist Girth (cm)") +
    ggtitle("Comparing Weight and Waist Girth for Men and Women") +
      theme(plot.title = element_text(hjust = 0.5))

```


There is convincing evidence that the Separate Lines Model is not as good of a fit as the Separate Lines Quadratic model (Sum of Squares F-Test. p-value = 0.0041). Having a quadratic term in this model makes sense since the right tail does not appear to be as linear as the rest of the terms. Statistically, the model is a better fit since it closer maps to the data. Thhe introduction of a quadratic term and its interactions show that the linear term for wai.gi is still significant but now the the interaction between wai.gi and MALE is no longer significant. Instead, the interaction between quadratic wai.gi and MALE is now significant. This indicates that the Waist Girth of males reduces weight in a quadratic fashion. The negative coefficient for the quadratic Waist Girth, MALE interaction mirrors that of the Separate Lines model meaning that the overall results are still the same but they can be predicted better using the quadratic model.