---
title: 'Homework #2'
author: "Dustin Leatherman"
date: "1/18/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(gridExtra)
library(forecast)
```

> Consider a signal-plus-noise model of the general for $x_t = s_t + w_t$ for t: [1, 100] where
> $s_t = \begin{cases} 0 & t:[1, 100]\\ 10 \ exp(- \frac{t - 100}{20}) cos(\frac{2 \pi t}{4}) & t: [101, 200]\end{cases}$
>
> and $w_t$ is Gaussian White Noise with $\sigma_w^2 = 1$
> 
> a) Simulate (set.seed(123)) and plot n = 200 observations from the model
> b) Although the model is not stationary, the ACF can be informative. For the data you generated, plot the ACF then comment

# a

```{r fig.width=10}
set.seed(123)

generateSignal <- function(n, cutoff) {
  # assume standard gaussian white noise
  w <- rnorm(n)
  
  # generate a signal (s)
  s <- function(t, cutoff) {
    ifelse(t <= cutoff, 0, 10 * exp(- (t - 100)/(20)) * cos((2 * pi * t)/(4)))
  }
  
  # combine white noise and signal
  w + sapply(seq(1, n), function(x) s(x, cutoff))
}

obs <- generateSignal(200, 100)


obs %>% 
  enframe %>% 
  ggplot(aes(x = name, y = value)) + 
    geom_point() + 
    geom_hline(yintercept = 0, linetype = "dashed") + 
    geom_line() +
    ylab(expression(x[t])) + 
    xlab("t")
```


# b

```{r fig.width=10}
autoplot(acf(obs, type = "correlation", plot = FALSE))
```

As the comparison between the current value and the number of lags increases, correlation slightly decreases yet remains significant. This is typically indicative of a trend within the data. The correlation value oscilates between positive and negative values for every other lag indicating that the dependence between the current value and the subsequent lags is _negative_. This can be confirmed by the scatterplot which shows choppiness from t = 100, ..., 140.

Additionally, even-numbered lags are significantly correlated with the current value of the time series. This may indicate that there is consistent interference with the data based on a multiple of 2. This makes sense since with know $cos(\frac{2 \pi t}{4})$ is a component of the signal and thus its frequency being $\frac{1}{4}$.