---
title: 'Homework #5'
author: "Dustin Leatherman"
date: "October 20, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(coin)
library(ggplot2)
library(dplyr)
library(broom)
library(grid)
library(gridExtra)
library(knitr)

tips$percent <- (tips$tip / tips$total_bill) * 100
attach(tips)

# hardcoding this because the problem states that there are two outliers and there are technically more.
tips_sans_outliers <- subset(tips, percent < 40)
```

# 1. Tip Percentage and Gender

#### Student t-test

``` {r summary, echo = TRUE, fig.width=10}

hist_full <- ggplot(tips, aes(x = percent, fill = sex)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~sex, ncol = 1)

bplot_full <- ggplot(tips, aes(x = sex, y = percent , fill = sex)) + 
  geom_boxplot(show.legend = FALSE)

grid.arrange(bplot_full, hist_full,
             ncol = 2, 
             widths = c(1, 2), 
             top = textGrob("Tip Percentage by Gender",
                            gp=gpar(fontsize=14,font=1),just=c("center")))
full_test <- t.test(percent ~ sex, data = tips)

hist_sans <- ggplot(tips_sans_outliers, aes(x = percent, fill = sex)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~sex, ncol = 1)

bplot_sans <- ggplot(tips_sans_outliers, aes(x = sex, y = percent , fill = sex)) + 
  geom_boxplot(show.legend = FALSE)

grid.arrange(bplot_sans, hist_sans,
             ncol = 2, 
             widths = c(1, 2), 
             top = textGrob("Tip Percentage by Gender (Outliers excluded)",
                            gp=gpar(fontsize=14,font=1),just=c("center")))

sans_test <- t.test(percent ~ sex, data = tips_sans_outliers)

# Put both tests in the same graph
knitr::kable(
  bind_rows(tidy(full_test) %>% 
              mutate(desc="With Outliers"), 
            tidy(sans_test) %>% 
              mutate(desc="Without Outliers")
            ) %>% 
    select(desc,estimate1, estimate2, p.value, conf.low, conf.high) %>% 
    mutate(estimate=estimate1 - estimate2), 
  digits = 4, 
  col.names = c("","Mean Female %", "Mean Male %", "P-Value", "C.I - Lower", "C.I - Upper", "Estimate"), 
  align = 'l')

```

The p-values produced by the full and sans outliers datasets are 0.2542 and 0.1336 respectively. The difference between these p-values is 0.1206.

#### Wilcoxon Rank Sum

``` {r rankSum, echo = TRUE, fig.width=10}
wilcox_test(percent~sex, data=tips, distribution="exact")
wilcox_test(percent~sex, data=tips_sans_outliers, distribution="exact")

```

The p-values produced by the full and sans outliers datasets are 0.1332 and 0.1479 respectively. The difference between these p-values is 0.0147.

The p-value difference produced by the Wilcoxon Rank Sum Test are significantly smaller than the p-value difference produced by the t-test. This suggests that the Wilcoxon Rank Sum Test is more robust to outliers than the t-test.

# 2. Plant Growth

``` {r plantGrowth, echo = TRUE, fig.width=10}
plant_bplot <- ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) + 
  geom_boxplot(show.legend = FALSE)

plant_hist <- ggplot(PlantGrowth, aes(x = weight, fill = group)) + 
  geom_histogram() +
  facet_wrap(~group, ncol = 1)


grid.arrange(plant_bplot, plant_hist,
             ncol = 2,
             widths = c(1, 2), 
             top = textGrob("Dried Plant Weights by Treatment (kg)",
                            gp=gpar(fontsize=14,font=1),just=c("center")))

knitr::kable(
  PlantGrowth %>%
    group_by(group) %>%
    summarise_all(funs(mean, sd, length, min, "25%"=quantile(weight, 0.25), median, "75%"=quantile(weight, 0.75), max)),
  digits = 4,
  col.names = c("Treatment Group","Mean","Std. Dev","Sample Size", "Min", "Q1", "Median", "Q3", "Max"), 
  align = 'l'
)
```


``` {r pooledInfo, echo = TRUE}

# Calculate pooled standard deviation, and RSS for full and reduced models
sums <- PlantGrowth %>% 
  group_by(group) %>% 
  summarise_all(funs(
    avg=mean, 
    stdev=sd, 
    n=n(), 
    ss.full = (weight - avg)^2 %>% sum, 
    ss.reduced = (weight - mean(PlantGrowth$weight))^2 %>% 
      sum)
    ) %>% 
  mutate(
    pool.stdev = sqrt(sum((n - 1) * stdev^2) / sum(n - 1)), 
    rss.full = sum(ss.full), 
    rss.reduced = sum(ss.reduced)
    )

kable(sums,
      digits = 4)

# Confirm that RSS for full and reduced models were calculate correctly
kable(sums %>% 
  select(rss.full, pool.stdev, rss.reduced, n) %>% 
  distinct %>% 
  mutate(
    rss.full.conf = (n*3 - 3) * pool.stdev^2,
    rss.reduced.conf = (n*3 - 1) * sd(PlantGrowth$weight)^2,
    rss.full.equal = assertthat::are_equal(rss.full, rss.full.conf),
    rss.reduced.equal = assertthat::are_equal(rss.reduced, rss.reduced.conf)),
  digits = 4
)
```

## ANOVA Table

|| Sum of Squares | df | MSS | F-value | p-value |
| -------- | -------- | ---------- | --------- | -------- | -------- |
| **Full** | 3.7663   | 2          | 1.8832    | 4.8461   | 0.0159   |
| **Reduced** | 10.4921 | 27       | 0.3886    |          |          |
| **Total** | 14.2584 | 29         |           |          |          |

There is moderate evidence that the average weight between each group in the study is different (extra sum of squares F-test on 27 and 2 degrees of freedom. p-value = 0.0159)

```{r tTest, echo = TRUE}
results <- t.test(weight~group, data = PlantGrowth[PlantGrowth$group != "trt2",])

kable(
  tidy(results) %>%
    select(estimate1, estimate2, p.value, conf.low, conf.high) %>% 
    mutate(estimate=estimate1 - estimate2),
    digits = 4, 
  col.names = c("Mean Weight Control", "Mean Weight Treatment 1", "P-Value", "C.I - Lower", "C.I - Upper", "Estimate"), 
  align = 'l')
```

With 95% confidence, the average weight difference between the control group and treatment group 1 is between -0.288 and 1.03 kg; thus there is no evidence to suggest that the average weights differ between the control and treatment group 1 (two-tail t-test. p-value = 0.2504).