---
title: 'Homework #7'
author: "Dustin Leatherman"
date: "November 1, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6)

library(tidyverse)
library(rjags)
library(knitr)
library(kableExtra)

```

> Let $Y_i$ be the precipitation for observation $i$ and $X_{ij}$ equal one if OTU j is present in sample $i$.  First, extract the 50 OTU with the largest absolute correlation between $X_{ij}$ and $Y_i$. Then fit a Bayesian linear regression model with precipitation as the response and with these 50 covariates (and an intercept term) using three priors:

> 1. Uninformative Normal Priors: $\beta_j \sim N(0, 100^2)$
>
> 2. Hierarchical Normal Priors: $\beta_j | \tau \sim N(0, \tau^2), \ \tau^2 \sim InvGamma(0.01, 0.01)$
>
> 3. Bayesian LASSO: $\beta_j | \tau^2 \sim DE (0, \tau^2), \ \tau^2 \sim InvGamma(0.01, 0.01)$

> Compare convergence and the posterior distribution of the regression coefficients under these three priors. In particular, are the same OTU’s significant in all three fits?

```{r}
load("~/StatisticsMasters/bayesian/homes.rdata")

# Join OTU dataset with Homes metadata 
sorted_data <- 
  # Indicator if OTU is present
  as.data.frame(ifelse(OTU > 0, 1, 0)) %>% 
  rownames_to_column(var = "ID") %>% 
  # data cleansing
  mutate(
    ID = as.integer(str_remove(ID, ".I"))
  ) %>% 
  # join with metadata and sort by ids
  inner_join(homes, by = "ID") %>% 
  arrange(ID)


X <- sorted_data %>% select(starts_with("OTU")) %>% as.matrix()
Y <- sorted_data %>% select(MeanAnnualPrecipitation) %>% as.matrix()

# calculate top 50 OTUs by absolute correlation with rainfall
top_otu <- 
  as.data.frame(cor(X, Y)) %>% 
    rownames_to_column(var = "OTU") %>% 
    rename(corr = MeanAnnualPrecipitation) %>% 
    arrange(desc(abs(corr))) %>% 
    head(n = 50) %>% 
    select("OTU")

# Put the data in JAGS format
X <- 
  sorted_data %>% 
    #mutate(intercept = 0) %>% 
    #select(intercept, top_otu$OTU) %>% 
    select(top_otu$OTU) %>% 
    as.matrix()
    

Y <- sorted_data %>% select(MeanAnnualPrecipitation)
Y <- Y$MeanAnnualPrecipitation

# utility function to union the output of JAGS
unionJagsOutput <- function(jags_data, names) {
  colnames(jags_data[[1]]) <- names
  colnames(jags_data[[2]]) <- names
  return (
    jags_data[[1]] %>% 
      as_tibble() %>% 
      mutate(chain = "1", row_num = row_number()) %>%
      pivot_longer(starts_with("OTU"), names_to = "var", values_to = "value") %>% 
      union_all(
        jags_data[[2]] %>% 
        as_tibble() %>% 
        mutate(chain = "2", row_num = row_number()) %>% 
        pivot_longer(starts_with("OTU"), names_to = "var", values_to = "value")
      )
  )
}

convergenceDiags <- function(jags_data, names) {
  p1 <- 
    unionJagsOutput(jags_data, names) %>%
      ggplot(aes(x = row_num, y = value, color = chain)) + 
        geom_line() + 
        facet_wrap(~var, ncol = 5) +
        labs(x = "Iteration", y = "Posterior Value") + 
        ggtitle("Trace of OTU Covariates")

  # Geweke Statistic
  p3 <- geweke.diag(jags_data)
  
  list(p1 = p1, p3 = p3)
}
```


# Fitting the Models

Let's fit our different Models. First, some setup though.


```{r}
n <- length(Y)
p <- ncol(X)

data <- list(Y=Y,X=X,n=n,p=p)
names <- colnames(X)

params <- c("beta")

# Settings (automatically calculates the number of iterations needed based on inputs)

nBurn <- 10000
nChains <- 2
nSave <- 4000
nThin <- 10
nIter <- ceiling((nSave*nThin)/nChains)
```



## Uninformative Normal Prior

$\beta_j \sim N(0, 100^2)$


```{r, fig.height=10}
model_string1 <- textConnection("model{
  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(alpha + inprod(X[i,], beta[]), taue)
  }
  # Priors
  for(j in 1:p){
    beta[j] ~ dnorm(0,0.01)
  }
  alpha ~ dnorm(0,0.01)
  taue ~ dgamma(0.1,0.1)
}")

model <- jags.model(model_string1,data=data,n.chains=nChains,quiet=TRUE)
update(model,burn=nBurn,progress.bar="none")
samples1 <- coda.samples(model,variable.names=params,thin=nThin,n.iter=nIter, progress.bar="none")
```


### Convergence

```{r, fig.width=10, fig.height=12}
convergenceDiags(samples1, names)

# gelmen-rubin statistics
gelman.diag(samples1)$psrf %>% 
      kable(
        caption = "Gelman-Rubin Statistics for convergence"
      ) %>% 
      kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")
```

According to the Gelman-Rubin diagnostics, all the coefficients converge. The Geweke statistics show  convergence and the Trace plots support this conclusion.


## Hierarchical Normal Prior

$\beta_j | \tau \sim N(0, \tau^2), \ \tau^2 \sim InvGamma(0.01, 0.01)$

```{r, fig.height=10}
model_string2 <- textConnection("model{
  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(alpha + inprod(X[i,], beta[]), taue)
  }
  # Priors
  for(j in 1:p){
    beta[j] ~ dnorm(0, taub)
  }
  alpha ~ dnorm(0,0.01)
  taue ~ dgamma(0.01,0.01)
  taub ~ dgamma(0.01,0.01)
}")

model <- jags.model(model_string2,data=data,n.chains=nChains,quiet=TRUE)
update(model,burn=nBurn,progress.bar="none")
samples2 <- coda.samples(model,variable.names=params,thin=nThin,n.iter=nIter, progress.bar="none")
```

### Convergence

```{r, fig.width=10, fig.height=12}
convergenceDiags(samples2, names)
gelman.diag(samples2)$psrf %>% 
      kable(
        caption = "Gelman-Rubin Statistics for convergence"
      ) %>% 
      kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")

```

According to the Gelman-Rubin diagnostics, all the coefficients converge. The Geweke statistics show  convergence and the Trace plots support this conclusion.

## Bayesian LASSO

$\beta_j | \tau^2 \sim DE (0, \tau^2), \ \tau^2 \sim InvGamma(0.01, 0.01)$

```{r, fig.height=10}
model_string3 <- textConnection("model{
  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(alpha + inprod(X[i,], beta[]), taue)
  }
  # Priors
  for(j in 1:p){
    beta[j] ~ ddexp(0, taub)
  }
  alpha ~ dnorm(0,0.01)
  taue ~ dgamma(0.01,0.01)
  taub ~ dgamma(0.01,0.01)
}")

model <- jags.model(model_string3,data=data,n.chains=nChains,quiet=TRUE)
update(model,burn=nBurn,progress.bar="none")
samples3 <- coda.samples(model,variable.names=params,thin=nThin,n.iter=nIter, progress.bar="none")
```


### Convergence

```{r, fig.width=10, fig.height=11}
convergenceDiags(samples3, names)
gelman.diag(samples3)$psrf %>% 
      kable(
        caption = "Gelman-Rubin Statistics for convergence"
      ) %>% 
      kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")

```

According to the Gelman-Rubin diagnostics, all the coefficients converge. There are a few covariates that don't show signs of convergence in the Geweke Statistics; however the Trace plots show that there does appear to be convergence.

## Plot Comparison

```{r, fig.width=10, fig.height=15}

getDensity <- function(jags_data, names) {
  return (
    unionJagsOutput(jags_data, names) %>% 
      group_by(var) %>% 
      summarise(den_x = density(value)$x, den_y = density(value)$y)
  )
}


d1 <- getDensity(samples1, names) %>% mutate(prior = "uninformative")
d2 <- getDensity(samples2, names) %>% mutate(prior = "Hierarchical Gaussian")
d3 <- getDensity(samples3, names) %>% mutate(prior = "BLASSO")

d_all <- union_all(d1, d2) %>% union_all(d3)

d_all %>% 
  ggplot(aes(x = den_x, y = den_y, color = prior)) + 
  geom_line() + 
  geom_vline(xintercept = 0) + 
  facet_wrap(~var, ncol = 5) +
  labs(x = expression(beta), y = "Posterior Density") +
  ggtitle("Comparing Posterior Densities across Priors")

```

There does not appear to be a big difference in the posterior distributions with different priors. Since the density lines have similar peaks, it does appear that these 50 OTUs are significant in all three fits.
