---
title: 'Homework #8'
author: "Dustin Leatherman"
date: "November 9, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(rjags)
library(kableExtra)

gambia <- read.csv("~/Downloads/gambia.csv") %>%  as_tibble()
```

> In this assignment, you will perform random slopes logistic regression in JAGS using the Gambia data described in gambia.csv.
> 
> Let $Y_i$ be the binary response for individual *i* testing positive for malaria, $\nu_i \in \{1, ..., 65 \}$ denote the village of the individual *i* and $X_i = 1$ if individual *i* regularly sleeps under a bed-net and $X_i = 0$ otherwise. Fit the model
> $$logit[P(Y_i = 1)] = \alpha_{\nu_i} + X_i * \beta_{\nu_i}$$
> where $\alpha$ and $\beta_j$ are the intercept and slope for village *j*.
> The priors (independent over village and with each other) are $\alpha \sim N(\mu_{\alpha}, \sigma_{\alpha}^2)$ and $\beta_j \sim N(\mu_b, \sigma_b^2)$. Choose uninformative priors for the hyper-parameters $\mu_{\alpha}, \mu_b, \sigma_{\alpha}^2, \sigma_b^2$
> In your report, address the following questions:
> 1. Scientifically, why might the effect of bed-net vary by village?
> 2. Did the MCMC algorithm converge?
> 3. Do you see evidence that the slopes and/or intercepts vary by village?
> 4. Which village has the largest intercept? Slope? Does this agree with the data in these villages?


# 1 

> Scientifically, why might the effect of bed-net vary by village?

The villages may have be in various degrees of forest cover which impacts the number of mosquitos present. If there are more mosquitos present, there is a higher probability that one slips through the bed net as a person enters or exits. 
```{r}
n <- dim(gambia)[1]

Y <- gambia$Y
X <- gambia$X
v <- gambia$v
distinct.v <- unique(v) 

data <- list(Y = Y, X = X, v = v, n = n, distinct.v = distinct.v)
params <- c("beta", "alpha")

# Settings (automatically calculates the number of iterations needed based on inputs)
nBurn <- 15000
nChains <- 3
nSave <- 4000
nThin <- 3
nIter <- ceiling((nSave*nThin)/nChains)

unionJagsOutput <- function(jags_data) {
  data <- NULL
  for(chain in 1:length(jags_data)) {
    new.data <- 
      jags_data[[chain]] %>% 
        as_tibble() %>% 
        mutate(chain = factor(chain), row_num = row_number()) %>%
        pivot_longer(-c(chain, row_num), names_to = "col", values_to = "value")
    
    if(is.null(data)) {
      data <- new.data
    } else {
      data <- union_all(data, new.data)
    }
  }
  return (data)
}
```

```{r}
model_string <- textConnection("model{
  # Likelihood
  for(i in 1:n){
    Y[i] ~ dbern(p[i])
    logit(p[i]) <- alpha[v[i]] + X[i] * beta[v[i]]
  }
  
  # Priors
  for(j in distinct.v){
    alpha[j] ~ dnorm(mu[1], tau[1])
    beta[j] ~ dnorm(mu[2], tau[2])
  }

  for(z in 1:2) {
    mu[z] ~ dnorm(0, 0.001)
    tau[z] ~ dgamma(0.01, 0.01)
  }
}")

model <- jags.model(model_string,data=data,n.chains=nChains, quiet = TRUE)
update(model,burn=nBurn,progress.bar="none")
model1.out <- coda.samples(model,variable.names=params,thin=nThin,n.iter=nIter, progress.bar = "none")
model1.data <- unionJagsOutput(model1.out)
```

```{r, fig.width=10, fig.height=10}
model1.data %>% 
  filter(str_detect(col, "alpha")) %>% 
    ggplot(aes(x = row_num, y = value, color = chain)) + 
      geom_line() + 
      facet_wrap(~col, ncol = 6, scales = "free") +
      labs(x = "Iteration", y = "Posterior Value", caption = "Simulating within-village variance", title = "Trace Plots for Alpha")

model1.data %>% 
  filter(str_detect(col, "beta")) %>% 
    ggplot(aes(x = row_num, y = value, color = chain)) + 
      geom_line() + 
      facet_wrap(~col, ncol = 6, scales = "free") +
      labs(x = "Iteration", y = "Posterior Value", caption = "Simulating the effect of Mosquito Nets on being tested positive for Malaria", title = "Trace Plots for Beta")
```

```{r, fig.width=10}
g.out <- geweke.diag(model1.out)

geweke.data <- NULL
for(chain in 1:nChains){
  new.data <- g.out[[chain]]$z %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "col") %>% 
    rename(value = ".") %>% 
    mutate(
      chain = factor(chain)
    )
  
  #colnames(new.data) <- c("col", "chain")
  if(is.null(geweke.data)) {
    geweke.data <- new.data  
  } else {
    geweke.data <- union_all(geweke.data, new.data)
  }
}

geweke.data <- 
  geweke.data %>%
    mutate(
      col = str_replace_all(col, "(\\[|,)", "_") %>% str_remove("\\]"),
      sim_row = str_split(col, "_", simplify = TRUE)[,2],
      group = str_split(col, "_", simplify = TRUE)[,1]
    )

geweke.data %>% 
  ggplot(aes(x = value, fill = chain)) + 
    geom_histogram() + 
    facet_wrap(~group) +
    geom_vline(xintercept = -2, linetype = "dashed") +
    geom_vline(xintercept = 2, linetype = "dashed") +
    scale_color_discrete(name = "Chain") +
    labs(x = "Geweke Statistic", y = "Density", title = "Geweke Distribution per Village", caption = "Histograms highlighting the convergence of the chains using the Geweke Statistic. It is ideal to be between the dotted lines.")
```


```{r}
gelman.out <- gelman.diag(model1.out)
gelman.out$psrf %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "col") %>% 
  as_tibble() %>% 
  summarise_at(c("Point est.", "Upper C.I."), list(quantile)) %>% 
  mutate(
    quantile = c("min", "0.25", "0.5", "0.75", "max")
  ) %>% 
  select(quantile, "Point est.", "Upper C.I.") %>% 
      kable(
        caption = "Gelman-Rubin Statistic Quantiles to measure convergence of chains"
      ) %>% 
      kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")

```

# 2

> Did the MCMC algorithm converge?

The Trace Plots, Geweke Distributions, and Gelman-Rubin Statistics indicate that the model has converged.

# 3

> Do you see evidence that slopes and/or intercepts vary by village?

```{r}
post.means <-
  model1.data %>% 
    group_by(col) %>% 
    summarise(post.mean = mean(value)) %>% 
    mutate(
        col = str_replace_all(col, "(\\[|,)", "_") %>% str_remove("\\]"),
        sim_row = str_split(col, "_", simplify = TRUE)[,2] %>% as.integer(),
        group = str_split(col, "_", simplify = TRUE)[,1]
      ) %>% 
    arrange(group, sim_row)

  
post.means %>% 
  filter(group == "alpha") %>% 
  summarize(value = quantile(post.mean, probs = c(0, 0.05, 0.25, 0.5, 0.75, 0.95, 1))) %>% 
  mutate(quantile = c("min", "5%", "25%", "median", "75%", "95%", "max")) %>% 
  select(quantile, value) %>% 
  kable(
    caption = "Quantiles of the Posterior Means for Intercepts"
  ) %>% 
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")

post.means %>% 
  filter(group == "beta") %>% 
  summarize(value = quantile(post.mean, probs = c(0, 0.05, 0.25, 0.5, 0.75, 0.95, 1))) %>% 
  mutate(quantile = c("min", "5%", "25%", "median", "75%", "95%", "max")) %>% 
  select(quantile, value) %>% 
  kable(
    caption = "Quantiles of the Posterior Means for Slopes"
  ) %>% 
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")
  
```

Slopes don't appear to vary much between village but intercepts range from -1.87 to 1.77, meaning that without taking into account bed nets, there is a predisposition to Malaria based on the village. This may indicate that there is information not being taken into account in this model.

# 4

> Which village has the largest intercept? Slope? Does this agree with the data in these villages?


```{r}
gambia %>% 
  group_by(v) %>% 
  summarise(infected = sum(Y), n = n(), rate = infected / n) %>% 
  arrange(desc(rate)) %>% 
  head(n = 10) %>% 
  kable(
    caption = "Infected Rates per village"
  ) %>% 
  kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position")
```

Village 49 has the largest intercept. The data shows that village 49 has the highest estimated effected rate via the MLE.
