---
title: 'Spatial modeling of gun-related homicide rates'
author: "Phillip Yates"
date: "10/29/2020"
output:
  pdf_document: default
  html_document: default
---

**Storing this here as a reference because it's a good example of Modeling for Spatial Correlation and it hasn't been covered in any class yet so I will want to look at this later.**

First we will load the `tidyverse` library for data reshaping and visualization.
```{r,message=FALSE}
library(tidyverse)
```

The data from this analysis come from "Firearm legislation and firearm morality in the USA: a cross-sectional, state-level study" by Kalesan et al. (2016). The response variable, $Y_{i}$, is the log firearm-related death rate (i.e., the log of the number of deaths divided by the population) in 2010 in state $i$. This is regressed onto five potential cofounders.

  1. log 2009 firearm death rate per 10,000 people
  2. Firearm ownership rate quartile
  3. Unemployment rate quartile
  4. Non-firearm homicide rate quartile
  5. Firearm export rate quartile
  
The covariate of interest is the number of gun control laws in effect in the state. This gives $p=6$ covariates.

We fit the linear model
$$Y_{i}=\beta_{0}+\sum^{p}_{j=1}X_{i}\beta_{j}+\epsilon_{i}.$$ We compute the usual non-spatial model with $\epsilon_{i}\sim\mbox{Normal}\left(0,\sigma^{2}\right)$ with the spatial model $\mbox{Cov}\left(\epsilon_{1},\ldots,\epsilon_{n}\right)\sim\mbox{Normal}(\mathbf{0},\pmb{\Sigma})$. The covariance
$$\pmb{\Sigma}=\tau^{2}\mathbf{S}+\sigma^{2}\mathbf{I}_{n}$$ is decomposed into a spatial covariance $\tau^{2}\mathbf{S}$ and a non-spatial covariance $\sigma^{2}\mathbf{I}_{n}$ where $\mathbf{I}_{n}$ is the $n\times n$ identity matrix. The spatial covariance follows the conditionally-autoregressive model $\mathbf{S}=\left(\mathbf{M}-\rho\mathbf{A}\right)^{-1}$, where $\mathbf{A}$ is the adjacency matrix with $(i,j)$ element equal 1 if states $i$ and $j$ are neighbors and zero otherwise, and $M$ is the diagonal matrix with $i^{th}$ diagonal element equal to the number of states that neighbor state $i$.

# Load the data

```{r}
load("~/StatisticsMasters/bayesian/guns.rdata")
Y <- log(10000*Y/N)
Z[,1] <- log(Z[,1]) # Z is also the 2nd column in X
X <- cbind(1,Z,rowSums(X))

# Remove Alaska and Hawaii
Y <- Y[-c(2,11)]
X <- X[-c(2,11),]

n <- length(Y)
p <- ncol(X)
```

# Fit the non-spatial model

```{r,message=FALSE}
ns_model <- "model{

  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(mu[i],taue)
    mu[i] <- inprod(X[i,],beta[])
  }
  # Priors
  for(j in 1:p){
    beta[j] ~ dnorm(0,0.01)
  }
  taue ~ dgamma(0.1,0.1)
  sig <- 1/sqrt(taue)
}"

library(rjags)
dat <- list(Y=Y,n=n,X=X,p=p)
init <- list(beta=rep(0,p))
model1 <- jags.model(textConnection(ns_model),inits=init,data=dat,quiet=TRUE)
update(model1,10000,progress.bar="none")
samp1 <- coda.samples(model1,variable.name=c("beta","sig"),n.iter=20000,
                      progress.bar="none")
summary(samp1)
```

# Create an adjacency matrix for the states in the US

```{r,message=FALSE}
library(maps)
library(spdep)
library(maptools)
usa.state <- map(database="state",fill=TRUE,plot=FALSE)
# The strsplit function splits the elements of a character vector into substrings
state.ID <- sapply(strsplit(usa.state$names,":"),function(x) x[1])
usa.poly <- map2SpatialPolygons(usa.state,IDs=state.ID)
# The poly2nb function builds a neighbors list from a "polygon" list
usa.nb <- poly2nb(usa.poly)
# The nb2mat function builds a spatial weights matrix for neighbors list
A <- nb2mat(usa.nb,style="B")
A <- A[-8,] # Take out DC
A <- A[,-8]
M <- diag(rowSums(A))
```

# Fit the spatial model

```{r}
sp_model <- "model{
  # Likelihood
    for(i in 1:n){
      Y[i] ~ dnorm(mu[i]+S[i],taue)
    }
    S[1:n] ~ dmnorm(zero[1:n],taus*Omega[1:n,1:n])
    for(i in 1:n){
      mu[i] <- inprod(X[i,],beta[])
      zero[i] <- 0
    }
    Omega[1:n,1:n] <- M[1:n,1:n]-rho*A[1:n,1:n]
    
  # Priors
    for(j in 1:p){
      beta[j] ~ dnorm(0,0.01)
    }
    taue ~ dgamma(0.1,0.1)
    taus ~ dgamma(0.1,0.1)
    rho ~ dunif(0,1)
    sig[1] <- 1/sqrt(taue)
    sig[2] <- 1/sqrt(taus)
}"

dat <- list(Y=Y,n=n,X=X,A=A,M=M,p=p)
init <- list(rho=0.99,beta=lm(Y~X-1)$coef)
model2 <- jags.model(textConnection(sp_model),inits=init,data=dat,quiet=TRUE)
update(model2,10000,progress.bar="none")
samp2 <- coda.samples(model2,variable.names=c("beta","rho","sig"),n.iter=20000,
                      progress.bar="none")
summary(samp2)
```

```{r}
rho <- samp2[[1]][,8]
ggplot(NULL,aes(x=rho))+geom_histogram(bins=16,color="white")+
  labs(x=expression(rho),y="Count")
```

**Summary:** The spatial dependence parameter is estimated to be near 0.35, indicating weak spatial dependence

# Compare the results across models

The objective is to determine if the coefficient corresponding to the number of gun laws, $\beta_{7}$, is non-zero. Below we compare its posterior distribution for the spatial and non-spatial models.

```{r}
b1 <- samp1[[1]][,7]
b2 <- samp2[[1]][,7]
d1 <- density(b1)
d2 <- density(b2)
Model <- c(rep("Non-spatial",length(d1$x)),rep("Spatial",length(d2$x)))
x <- c(d1$x,d2$x)
y <- c(d1$y,d2$y)
density.data <- as.data.frame(x=x,y=y,Model=Model)
ggplot(density.data,aes(x=x,y=y,color=Model))+geom_line()+
  labs(x=expression(beta[7]),y="Posterior Density")
```

```{r}
mean(b1<0)
mean(b2<0)
```

**Summary:** Both models provide evidence of a negative relationship between the number of gun laws and the firearm-related death rate. However, there is more uncertainty in the spatial model, which is likely more realistic.