---
title: 'Data Analysis #2'
author: "Dustin Leatherman"
date: "5/24/2019"
output:
  pdf_document: default
  html_document: default
always_allow_html: yes
bibliography: LeathermanDataAnalysis2.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE, fig.width = 10)

library(MASS)
library(proj4)
library(leaflet)
library(ggplot2)
library(grid)
library(gridExtra)
library(tidyverse)
library(reshape2)
library(knitr)
library(kableExtra)
library(GGally)
library(pscl)


dind <- function (glm.full, glm.reduced) {
  lrt <- glm.reduced$deviance - glm.full$deviance
  lrt.df <- glm.reduced$df.residual - glm.full$df.residual
  1 - pchisq(lrt, lrt.df)
}
```

# Introduction

It is common practice for industries that harvest natural resources, such as logging and mining, to replace trees after they have been removed. The impact of climate change on adolescent trees is not well understood.  South Western Australia provides a unique test bed for assessing climate change impact since it is one of the few places that have already felt significant climate change. Mean rainfall has experienced a 17% reduction between 1975-2011 compared to the mean rainfall between 1900-1974. In parallel, the average daily highs and lows have been increasing which exacerbates droughts [@JarrahForest; @IndianOcean]. 

Jarrah trees (*Eucalyptus marginata*) are canopy trees in South Western Australia that thrive in the 650-1250mm rainfall zone [@Jarrah]. Jarrah are resistant to termites and pests but particularly susceptible to Dieback or root rot. Dieback is caused by bacteria that travels in water and is absorbed through a root system [@Diebeck]. One method in identifying wet and dry season is by using an Ombrothermic diagram. It summarizes trends in temperature and precipitation to allow for establishing a relationship between the two. The length of the wet season is determined by ombrothermic relationship where average rainfall exceeded average high temperatures [@Ombrothermic]. The dataset contains aggregate data so rebuilding ombrothermic diagrams for each year is not possible but it is still a valuable tool in understanding how the data was calculated.

## Questions of Interest

* What is the effect of temperature, rain amount, rain consistency, and diebeck on the species richness of plot-level vegetation?
* Is the length of a wet season a factor in whether soil is afflicted with Diebeck?

\pagebreak

# Sample Collection

Seedlings were established in 1938 plots total between the restoration sites of the Huntly and Willowdale mines. The seedlings were monitored 15 months after planting as well as after the onset of the first wet season from 1992-2010. Data were recorded as species frequencies ranging from 0-100 based on their presence within five 4 x 4m quadrats nested within 20 x 20m plots. The data describes the species composition of the restored Jarrah fores, including 491 plant species in total [@JarrahForest].

The Climate variables were measured from climate stations at the Huntly and Willowdale mining sites while additional air temperature statistics were taken from Dwellingup, the meteorological station nearest to both sites. Not captured in this particular dataset is historical data the researchers used to determine the surety of the effect of climate change in the area. Mediterranean climates are defined by long summer droughts when rainfall is less than twice the mean temperature. This relationship was used to define the wet season for each year [@JarrahForest].

73.5% of the plots recorded a diebeck status so those will be the only ones used in analysis. This is a potential gap for accurately predicting diebeck.


```{r map, fig.height=8}
climate <- read.csv("~/Downloads/Standish et al. JEcol-2014-0252R1.csv")
# friendly name for legends
climate$MineFull <- ifelse(climate$Mine == "HU", "Huntly", "Willowdale")

# x,y are required for project() to convert UTM coords to lat-long
projDf <- data.frame(x = climate$Easting, y = climate$Northing)

# UTM Coordinates for south western Australia where the study took place
proj4Config <- "+proj=utm +zone=50 +south +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
pj <- project(projDf, proj4Config, inverse = TRUE)
climate$lat <- pj$y
climate$lon <- pj$x

# Make a pretty Topographical map. colored by mines
pal <- colorFactor(c("red", "blue"), domain = c("Huntly","Willowdale"))
leaflet(data = climate, options = leafletOptions(zoomControl = FALSE)) %>% 
  addProviderTiles(providers$OpenTopoMap) %>%
  addCircleMarkers(
    radius = 3,
    color = ~pal(MineFull),
    stroke = FALSE
  ) %>%
  addLegend("bottomright", pal = pal, values = ~MineFull,
            title = "Mine",
            opacity = 1
  )
```

\pagebreak

# Summary Statistics

```{r summaryTables}
climateCalcs <- 
  climate %>% 
    select_if(is.numeric) %>% 
    select(-Easting, -Northing, -Elevation, -Slope, -Year) %>%
    summarize_all(
      funs(
        Min=min, 
        Q25 = quantile(., 0.25, na.rm = TRUE), 
        Median=median, 
        Q75 = quantile(., 0.75, na.rm = TRUE), 
        Max=max, 
        Mean=mean, 
        Stdev = sd
        )
      ) %>% 
    gather(stat, val) %>%
    separate(stat, into = c("Variable", "stat"), sep = "_") %>%
    spread(stat, val) %>%
    column_to_rownames("Variable") %>%
    select(Min, Q25, Median, Q75, Max, Mean, Stdev)

climateCalcs %>%
  kable(
    digits = 4,
    caption = "Jarrah Restoration for 1992-2010 (n = 1938). Bolded rows indicate variables used in subsequent analysis",
    row.names = TRUE
  ) %>% 
    kable_styling(full_width = T, bootstrap_options = "striped", latex_options = "hold_position") %>%
    row_spec(which(rownames(climateCalcs) %in% c("Rain30", "Rain60", "Rain90", "Rain120", "RainWS", "Rain15mth", "RainEven30", "RainEven30", "RainEven60", "RainEven90", "RainEven120", "RainEvenWS", "RainEven15mth", "HighestTemp", "LowestTemp", "AvgMaxTemp", "AvgMinTemp", "Rain.Length", "Veg.SR")), bold = T)
```

```{r summaryDefs}

# Averages and Extreme
plot.temps <-
  climate %>% 
  select(Year, `Average Low` = AvgMinTemp, `Average High` = AvgMaxTemp, `High` = HighestTemp, `Low` = LowestTemp) %>% 
  gather(c("Average High", "High", "Low", "Average Low"), key = "Temperature", value = "value") %>%
  mutate(variable = factor(Temperature, level = c("High", "Average High", "Average Low", "Low"))) %>%
  ggplot(aes(x = Year, color = Temperature, y = value)) + 
    geom_line() +
    xlab("Year") + ylab("Temperature (C\u00B0)") +
    labs(
      title = "Estimated Annual Temperature",
      caption = "Homogenous results across both sites"
    )

climate.groups <-
  climate %>% 
  select(Year, Mine=MineFull, `30 Days` = Rain30, `60 Days` = Rain60, `90 Days` = Rain90, `120 Days` = Rain120, `15 Months` = Rain15mth, `Wet Season Total` = RainWS) %>% 
  distinct %>% 
  gather(c(-Year,-Mine), key = "variable", value = "value") %>% 
  # reorder factors for presentation purposes
  mutate(variable = factor(variable, levels = c("30 Days", "60 Days", "90 Days", "120 Days", "15 Months", "Wet Season Total")))

climate.groups.avg <- 
  climate.groups %>%
  group_by(Mine, variable) %>%
  summarize_at("value", funs(avg=mean))

# Line charts for Rainfall
plot.scatter.rainfall <-
  climate.groups %>% inner_join(climate.groups.avg, by = c("Mine","variable")) %>%
  ggplot(aes(x = Year, y = value, color = Mine)) + 
    geom_line() +
    geom_line(aes(y = avg), linetype = "dashed") +
    facet_wrap(~ variable , ncol = 3) +
    xlab("Year") + ylab("Rainfall (mm)") +
    labs(
      title = "Rainfall Throughout the Wet Season",
      caption = "Dashed lines indicate average"
    )

climate.raineven <-
  climate %>% 
  select(Year, Mine=MineFull, `30 Days` = RainEven30, `60 Days` = RainEven60, `90 Days` = RainEven90, `120 Days` = RainEven120, `15 Months` = RainEven15mth, `Wet Season Total` = RainEvenWS) %>% 
  gather(c(-Year,-Mine), key = "variable", value = "value") %>% 
  # reorder factors for presentation purposes
  mutate(variable = factor(variable, levels = c("30 Days", "60 Days", "90 Days", "120 Days", "15 Months", "Wet Season Total")))

climate.raineven.avg <-
  climate.raineven %>% 
  group_by(Mine, variable) %>%
  summarize_at("value", funs(avg=mean))

# Line charts that show Rain Evenness
plot.scatter.raineveness <-
  climate.raineven %>% inner_join(climate.raineven.avg, by = c("Mine","variable")) %>%
  ggplot(aes(x = Year, y = value, color = Mine)) + 
    geom_line() + 
    geom_line(aes(y = avg), linetype = "dashed") +
    facet_wrap(~ variable , ncol = 3) +
    xlab("Year") + ylab("Shannon Diversity Index") +
    labs(
      title = "Rainfall Evenness Throughout the Wet Season",
      caption = "0 = complete unevenness. 1 = complete evenness. Dashed lines indicate average"
    )

climate.wet.season <-
  climate %>% 
    select(Rain.Length, Year, Mine=MineFull) %>% 
    group_by(Mine, Year) %>% 
    distinct

climate.wet.season.mine <-
  climate.wet.season %>% 
    group_by(Mine) %>%
    summarise_at("Rain.Length", funs(avg = mean))

plot.wetseason <-
  climate.wet.season %>% 
  inner_join(climate.wet.season.mine, by = "Mine") %>%
  ggplot(aes(x = Year, y = Rain.Length, color = Mine)) +
    geom_line(show.legend = FALSE) +
    geom_line(aes(y = avg), linetype = "dashed", show.legend = FALSE) +
    xlab("Year") + ylab("Days") +
    labs(
      title = "Length of Wet Season",
      caption = "Dashed lines indicate average"
    )

diebeck <- 
  climate  %>% 
  drop_na(Dieback) %>% 
  mutate(isDieback = ifelse(Dieback == "DBF", 1, 0))

diebeck.by.year <- 
  diebeck %>% 
  group_by(MineFull) %>% 
  summarize_at("Year", funs(cnt = n()))

diebeck.by.year.tot <-
  diebeck %>%
  select(isDieback, Year, MineFull) %>% 
  group_by(MineFull, Year) %>% 
  summarise_at("Year", funs(cnt=n())) %>% 
  inner_join(diebeck.by.year, by = "MineFull") %>% 
  mutate(prcnt.tot = cnt.x / cnt.y)

diebeck.by.mine <-
  diebeck.by.year.tot %>% 
  group_by(MineFull) %>%
  summarize_at("prcnt.tot", funs(avg=mean))
# percent of isDiebeck
plot.diebeck <-
  diebeck.by.year.tot %>% 
  inner_join(diebeck.by.mine, by = "MineFull") %>%
  rename(Mine = MineFull) %>%
  ggplot(aes(x = Year, y = prcnt.tot, color = Mine)) + 
  geom_line() +
  geom_line(aes(y = avg), linetype = "dashed") +
  xlab("Year") + ylab("Percentage") +
  labs(
      title = "Trees Afflicted with Diebeck",
      caption = "Dashed lines indicate average"
    )

climate.veg.vars <- climate %>% select(Veg.SR, Year, Mine=MineFull)
climate.veg <-
  climate.veg.vars %>% 
    group_by(Mine, Year) %>% 
    summarize_at("Veg.SR", funs(avg = mean))

climate.veg.mine <-
    climate.veg.vars %>% 
    group_by(Mine) %>%
    summarise_at("Veg.SR", funs(avg = mean))
  
plot.veg <-
  climate.veg.vars %>% 
    inner_join(climate.veg, by = c("Year", "Mine")) %>% 
    inner_join(climate.veg.mine, by = "Mine") %>%
    ggplot(aes(x = Year, y = Veg.SR, color = Mine)) +
      geom_point(alpha = 0.6) +
      geom_line(aes(y = avg.x), show.legend = FALSE) +
      geom_line(aes(y = avg.y), linetype = "dashed", show.legend = FALSE) +
      xlab("Year") + ylab("Number of Species") +
      labs(
        title = "Number of Species Found",
        caption = "Dashed lines indicate average for all years. Solid lines indicate average across years"
      )
```

## A Closer Look

Though not pictured, Rainfall N is positively correlated with with Rainfall (N+30). Rainfall is generally more consistent at the Huntly mining site. At the start of the wet season, the consistency at both sites is variable year-over-year but tends to flatten out as the season progresses. This is evident from the converging averages as well as the less erratic nature of measures for the Wet Season Total. For simplicity and to resolve multicollinearity, Wet Season Totals will be selected since they are similar for both mine sites and contain the least amount of variation. As expected, rainfall increases throughout the wet season but oddly enough, rainfall after 15 months is much higher than the season total. This is a curiousity that would be explored more if this variable were selected.

```{r summary.rainfall}
# show the plots here. This is done so I can quickly change display order
plot.scatter.rainfall
plot.scatter.raineveness
```

```{r summary.climate}
grid.arrange(plot.wetseason, plot.diebeck, ncol = 2)
plot.temps
```

The length of wet season and the trees afflicted with Diebeck are variable year-over-year. For a more consistent measure, serial correlation may be present.

```{r summary.veg}
plot.veg
```

Willowdale in general appears to have a higher number of species found than Huntly. There also appear to be a few more large outliers at Willowdale compared to huntly which influence the overall average. The peaks and valleys of this chart do not seem to line up as nicely as the others indicating that perhaps climate variables are not the biggest influence in number of species found.

### Assessing Transformations

The strongest correlation is between Wet Season Length and Rainfall for the Wet Season with a coefficient of 0.718 across both campsites. There appears to be a log-like relationship between Average Min Temperature and Rain Evenness but given that Rain Evenness is an index, a log transformation does not affect its shape.

```{r corrs, fig.height=8}
climate %>% 
  select(Mine=MineFull, AvgMinTemp, AvgMaxTemp, HighestTemp, LowestTemp, RainWS, RainEvenWS, Rain.Length) %>%
  ggpairs(
    mapping = aes(color = Mine), 
    columns = c("AvgMinTemp", "AvgMaxTemp", "HighestTemp", "LowestTemp", "RainWS", "RainEvenWS", "Rain.Length"),
    legend = c(1,1)
  )

```

### 2002 & 2006

2002 is the maxima for percentage of trees affected with Diebeck for the Huntly mine. It also appears to be a maxima for rainfall, rainfall evenness for Wet Season Totals, and a minor maxima for the length of Wet Season. 2006 is the minima for percentage of trees affected with Diebeck and similar minimas are apparent in the aforementioned variables. This indicates that there may be a relationship between trees being affected Diebeck and rainfall. 2006 yields a maxima for the average high and minimas for High and Low temperatures. No such relationship is apparent for 2002. 


# Inferential Procedures

Species Richness is a measure describing the number of vegetation species found in a given plot. If each species is considered a success, then the number of species in a given plot can be considered some number of successes during a period of time; time in this case is measured by year. This can be modeled with by Negative Binomial or a Poisson General Linear Model. First, Poisson and Quasipoisson models will be reviewed for adequacy, followed by a Negative Binomial Model. A Likelihood Ratio Test between the Poisson and Negative Binomial models will be run in order to determine which one is more suitable.

Ascertaining whether or not a given plot is afflicted with diebeck can be done with a logistic regression model. Selecting mostly climate-related variables, a logistic regression model can be created and its error rates analyzed to determine whether or not Wet Season Length is a significant factor in determining whether or not a plot has diebeck. Since the data is time series and therefore not independent, a quasibinomial model will be used to account for overdispersion.

# Results

## Species

**Hypothesis**: It is predicted that an increase in average high temperatures and an increase in rain preciptation will yield more vegetation species in a plot. 

### Ruling out Poisson

Extra-poisson variation is likely since there are known impactful explanatory variables available in the dataset that are not being included in the model. Fitting the model and examining the deviance goodness-of-fit test in conjunction helps determine whether extra-poisson variation is present. All mean values produced by the model are greater than 5 meaning that it is also possible. There is convincing evidence that the poisson model is not inadequate (Deviance Goodness-of-fit Test. p-value = 0). Running the same test for the quasipoisson version model yields the same result (p-value = 0). This indicates that the Poisson distribution is an inadequate model for the response.

```{r quasipoisson}
model.species.poisson <- glm(Veg.SR ~ RainWS + RainEvenWS + AvgMaxTemp + AvgMinTemp + HighestTemp + LowestTemp + Rain.Length + isDieback , family = "poisson", data = diebeck)

gof4 <- 1 - pchisq(model.species.poisson$deviance, model.species.poisson$df.residual)

model.species.poissonq <- glm(Veg.SR ~ RainWS + RainEvenWS + AvgMaxTemp + AvgMinTemp + HighestTemp + LowestTemp + Rain.Length + isDieback , family = "quasipoisson", data = diebeck)

gof5 <- 1 - pchisq(model.species.poissonq$deviance, model.species.poissonq$df.residual)

```

A Negative Binomial model will be reviewed as an alternative to the Quasi-poisson model.

### Negative Binomial

The main negative binomial model in question:

$$
  \hat{log(Veg.SR)} = \beta_0 + \beta_1RainWS + \beta_2RainEvenWS + \beta_3AvgMaxTemp + \beta_4AvgMinTemp + \beta_5HighestTemp 
$$

$$
  + \beta_6LowestTemp + \beta_7Rain.Length + \beta_8isDiebeck
$$

$$
  \hat{log(Veg.SR)} = 5.919 - 0.00001RainWS + 0.16RainEvenWS - 0.0867AvgMaxTemp + 0.0094AvgMinTemp
$$

$$
- 0.0033HighestTemp + 0.0203LowestTemp - 0.0003Rain.Length - 0.0359isDiebeck
$$

There is no evidence that this model is inadequate (Deviance Goodness-of-Fit Test. p-value = 0.3244). A Maximum Likelihood Ratio Test indicates that there is convincing evidence that a Negative Binomial GLM is a better fit than a Poisson GLM (p-value = 0.3244).

Testing to see if a model with interaction between each of the variables and the diebeck indicator yield no evidence that the interaction adds value (Drop-in-Deviance Test. p-value = 0.9991).

There is convincing evidence that log Number of Species is associated with Average Max Temp, Lowest Temp, and the Diebeck indicator (Two-tailed Wald Test. p-value = 2.31e-06, 0.0089, 0.0069 respectively). 

It is estimated that the average number of species in a plot is 1.09 times higher for every degree celcius that the average max temperature decreases after fixing all other variables. With 95% confidence, a one degree celcius decrease in Average Max Temperature increases the average number of species in a plot by between 1.0515 and 1.1312 times.

It is estimated that the average number of species in a plot is 1.02 times higher for every degree celcius that the Lowest temperature increases after fixing all other variables. With 95% confidence, a one degree celcius increase in the Lowest Temperature increases the average number of species in a plot by between 1.0052 and 1.0359 times.

It is estimated that the average number of species in a plot is 1.03 times lower when diebeck is present after fixing all other variables. With 95% confidence, the presence of Diebeck decreases the average number of species in a plot by between 1.0099 and 1.064 times.

There is no evidence that Rainfall or Rain Evenness for the Wet Season have any impact on the number of species in the plot (Two-tailed Wald Test. p-value = 0.8505, 0.6629 respectively).


```{r species.res}
model.species.nb <- glm.nb(Veg.SR ~ RainWS + RainEvenWS + AvgMaxTemp + AvgMinTemp + HighestTemp + LowestTemp + Rain.Length + isDieback, data = diebeck)
#odTest(model.species.nb)

# 0.3244
gof3 <- 1 - pchisq(model.species.nb$deviance, model.species.nb$df.residual)

model.species.nb.full <- glm.nb(Veg.SR ~ (RainWS + RainEvenWS + AvgMaxTemp + AvgMinTemp + HighestTemp + LowestTemp + Rain.Length) * isDieback, data = diebeck)

# 0.9991
#dind(model.species.nb.full, model.species.nb)
```


## Diebeck

**Hypothesis**: It is expected that a longer wet season will yield a higher chance of a plot becoming diebeck.

The model with the parameters described above is:

$$
  logit(diabeck) = \beta_0 + \beta_1RainWS -\beta_2RainEvenWS + \beta_3AvgMaxTemp - \beta_4AvgMinTemp - \beta_5HighestTemp
$$

$$
  + \beta_6LowestTemp + \beta_7RainLength
$$

There is convincing evidence that this model is inadequate (Deviance Goodness of Fit Test. p-value = 1.1102e-16).

Even though the model is inadequate, the Wet Season Length parameter can still be assessed for significance. There is no evidence that the log odds of diebeck are associated with the length of the Wet Season (Wald Logistic Regression on a Single Variable. p-value = 0.3227). Given that this inadequate model contains Wet Season Length and it is not significant, it is feasible that this is enough evidence to discount this relationship. Further analysis with more data would need to be done in order to gain more insight.


```{r res.diebeck}
model.diebeck <- glm(isDieback ~ RainWS + RainEvenWS + AvgMaxTemp + AvgMinTemp + HighestTemp + LowestTemp + Rain.Length, family = "quasibinomial", data = diebeck)

gof1 <- 1 - pchisq(model.diebeck$deviance, model.diebeck$df.residual)
```

# Conclusion

## Species

The results surrounding Species prediction turned out to be different than anticipated. There was a small but significant positive multiplicative effect on Species richness when Average High Temperatures decreased and Low Temperatures increased. This is counter to the hypothesis where it was posited that the number of species would increase with average high temperatures and precipitation. Interestingly enough, precipitation had no effect on species. This may be due to the temperature increasing past the threshold when the Jarrah trees best take root. This would explain why an increase in Lowest Temperature also had an associated effect on the number of species taking root.

## Diebeck

The hypothesis was that the length of the Wet Season would increase the chances of soil becoming diebeck. The chosen parameters were not sufficient to provide an adequate logistic regression model. For the inadequate model that was fit, Wet Season Length did not significantly affect the log odds of diebeck which could be construed as not being significant. This is a possibility but a tenuous one at best until further analysis is done with more data.

## Recommendations

The goal of this analysis was to understand the impact of climate on diebeck in soil and the ability for vegetation species to thrive in the face of climate change. The original paper concludes that climate had a significant but small associated impact to number of species compared to the impact of healthy restoration practices. While this analysis did not include any of the restoration-centric explanatory variables (barring the diebeck indicator), it was able to provide similar conclusions.

This dataset has a wealth of possibilities but it requires a decent understanding in climate and forestry in order to draw conclusions. I reccommend appropriate time series analysis techniques and further research on the proper environmental conditions and growth practices for Jarrah trees.

## Limitations

Time Series analysis is a must for proper analysis. The analysis provided in this report is interesting but the conclusions are weak since the modeling does not account for trends, seasonality, or spatial analysis between sites. The biggest limitation for this analysis was experience with the appropriate analysis techniques required to correctly analyze this data. Further research in Time Series Analysis is needed in order to adjust for serial correlation between Years appropriately. I would have used more of the restoration variables had I understood them better. The particular focus for this analysis was mostly on climate with a touch of restoration but that led to weak models due to climate only being part of the equation.

\pagebreak

# Appendix

## Variables

### Temporal variable

|Name|Description|Type|
|------------|----------------------------------------------|----|
|Year|1992 to 2010|continuous|	

### Spatial variables

|Name|Description|Type|
|---------------|----------------------------------------------|----|
|Mine|Huntly (HU) or Willowdale (WD)|categorical|
|Temp.Plot|Year of Monitoring, then plot number|identifier|
|Easting|In UTMs to 2 d.p.|continuous|
|Northing|In UTMs to 2 d.p.|continuous|
|Elevation|Elevation in metres asl to 2 d.p.|continuous|
|Slope|Slope in degrees to 4 d.p.|continuous|
|Aspect|Aspect in degrees to 4 d.p.|continuous|

### Climate variables

|Name|Description|Type|
|---------------|----------------------------------------------|----|
|AvgMinTemp|Annual estimate|continuous|
|AvgMaxTemp|Annual estimate|continuous|
|LowestTemp|Annual estimate|continuous|
|HighestTemp|Annual estimate|continuous|
|DaysOver35|Per year|continuous|
|Rain.Length|Wet seasonal length in days|continuous|
|Drought.Length|Length of first summer drought in days|continuous|
|False.Start|1 = yes. 0 = no.|factor|
|Rain30|Rainfall amount 30 days after onset of wet season|continuous|
|Rain60|Rainfall amount 60 days after onset of wet season|continuous|
|Rain90|Rainfall amount 90 days after onset of wet season|continuous|
|Rain120|Rainfall amount 120 days after onset of wet season|continuous|
|RainWS|Wet season total rainfall|continuous|
|RainPrevWS|Previous year's wet season total rainfall|continuous|
|RainEven30|Rainfall evenness at 30 days\*|continuous|
|RainEven60|Rainfall evenness at 60 days\*|continuous|
|RainEven90|Rainfall evenness at 90 days\*|continuous|
|RainEven120|Rainfall evenness at 120 days\*|continuous|
|RainEvenWS|Rainfall evenness at end of wet season\*|continuous|
|RainEven15mth|Rainfall evenness at 15 months|continuous|
|Rain15mth|Total rainfall between onset of wet season and approximate date of vegetation monitoring|continuous|

\* Rainfall Evenness is calculated using the modified Shannon diversity index. Values vary from 0 to 1 where 0 implies complete unevenness (i.e. all rain recorded for the period falls in one event), and 1 implies complete evenness (i.e. rain recorded for the period falls in equal amounts each day) [@JarrahForest].

*(all data sourced from Bureau of Meterology, Dwellingup climate station, the closest station to HU and WD)*

### Variables describing restoration practice
|Name|Description|Type|
|---------------|----------------------------------------------|----|
Topsoil.rateALL|Ordinal scale for topsoil rtn|ordinal
StripSpread.delay|Delay between stripping the topsoil from the donor site and spreading it onto the restoration site: days.|continuous|
Wet.start|Date of the start of the wet season according to Ombrothermic relationship.|continuous|
SeedMix.tt|Seed mix treatment according to year of application and mine site|categorical|
SeedMix.SR|Number of species in the seed mix per year of application and mine site|continuous|
Dieback|Dieback-affected topsoil (DB) or dieback-free topsoil (DBF); NA = missing data|categorical|
FertP|Expressed kg elemental per ha|continuous|
FertN|Expressed kg elemental per ha|continuous|
FertK|Expressed kg elemental per ha|continuous|
SeedWetstart.syn|Number of days between seeding and the start of the wet season. Can be negative or positive|continuous|

### Response Variables
|Name|Description|Type|
|---------------|----------------------------------------------|----|
Similarity|Bray-Curtis similarity between plot-level vegetation and seed mix applied at HU or WD for relevant year (n= 518 species)|continuous|
Veg.SR|Species richness of plot-level vegetation (n= 491 species)|continuous|
Prop.Sp.shared|Sp.shared.with.Speciose' divided by 'Veg.SR'|continuous|

### Topsoil.rateALL

|Score|Description|
|---------------|----------------------------------------------|
Score 1|Fallow-stockpiled topsoil (F-STP); (Fresh)-stockpiled topsoil (STP); Fallow-sieved topsoil (F-SS); Stockpiled sieved topsoil (STPSS) and fallow stockpiled and screened topsoil (F-STPSS).|
Score 2|Fallow-direct returned topsoil, includes all ratios of return rates (F-DRT).|
Score 3|Fresh sieved topsoil at rate of 20 m2 (SS)|
Score 4|Fresh sieved topsoil at rate of 30 m2 (SS)|
Score 5|Fresh sieved topsoil at rate of 40 m2 (SS)|
Score 6|Fresh direct-returned 1: 4 unsieved topsoil; includes all return ratios greater than 1: 4 too (DRT)|
Score 7|Fresh direct-returned 1: 1 unsieved topsoil, 1: 2 unsieved topsoil, 1: 3 unsieved topsoil (DRT)|

## References
