---
title: 'Quiz #2'
author: "Dustin Leatherman"
date: "May 5, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
```

# 1

The following table is from a report on the relationship between aspirin use and myocardial infarction (heart
attacks) by the Physicians’ Health Study Research Group at Harvard Medical School.

### Myocardial Infarction

| Group   | Yes | No    | Total |
| ------- | --- | ----- | ----- |
| Placebo | 189 | 10845 | 11034 | 
| Aspirin | 104 | 10933 | 11037 |

The Physicians’ Health Study was a five-year randomized study testing whether regular intake of aspirin
reduces mortality from cardiovascular disease. Every other day, the male physicians participating in the
study took either one aspirin tablet or a placebo. They study was “blind” – the physicians in the study did
not know which type of pill they were taking.

## a 

*Compute and interpret a 95% confidence interval for the difference in proportion of myocardial infarction between the two treatments.*

$$
\hat{\pi_2} - \hat{\pi_1} \pm z_{0.975}\sqrt{\frac{\hat{\pi_1}(1 - \hat{\pi_2})}{n_1} + \frac{\hat{\pi_2}(1 - \hat{\pi_1})}{n_2}}
$$

$$
\frac{104}{11037} - \frac{189}{11034}  \pm 1.96\sqrt{\frac{0.0169(1 - 0.0095)}{11034} + \frac{0.0095(1 - 0.0169)}{11037}}
$$

UCL = -0.0047
LCL = -0.0107

With 95% confidence, Patients who took Aspirin were between 0.47% and 1.07% less likely to have a heart attack than Patients who took a Placebo.


## b

*Compute and interpret a 95% confidence interval for the odds ratio comparing the risk of myocardial infarction between the placebo and aspirin treatment groups.*

```{r 1b}
theta <- (189 * 10933) / (104 * 10845)
n1 <- 11034
n2 <- 11037
pi1 <- 189 / 11034
pi2 <- 104 / 11037
ucl <- log(theta) + qnorm(0.975) * sqrt((1 / (n1 * pi1 * (1 - pi1))) + (1 / (n2 * pi2 * (1 - pi2))))
lcl <- log(theta) - qnorm(0.975) * sqrt((1 / (n1 * pi1 * (1 - pi1))) + (1 / (n2 * pi2 * (1 - pi2))))

c(exp(lcl), exp(ucl))
```

With 95% confidence, Patients who took took the Placebo were between 1.44 and 2.3308 times more likely to receive a heart attack than those who took Aspirin.

# 2

*Is there a gender gap in political affiliation? The following table uses data from the 2000 General Social Survey. It cross classifies gender and political party identification more strongly with the Democratic or Republican party or as Independents.*

**<Table Not Shown>**

*Perform a test on whether gender and party identification are independent factors. State the null and alternative hypotheses, the test statistic, the p-value, and the conclusion. Do not forget to check the assumptions of the test.*

The appropriate test for checking for independence for an R x C table between Gender and Party Identification is the Chi-Squared Mantel-Haentzsel Test.

```{r 2}
# short answer
parties <- data.frame(Democrat=c(762,484), Independent=c(327,239), Republican=c(468,477), row.names = c("Females", "Males"))
chisq.test(parties)$expected %>% kable
chisq.test(parties)

# long answer
r1c1 <- (762 - ((1246 * 1557) / 2757))^2 / ((1246 * 1557) / 2757)
r1c2 <- (327 - ((566 * 1557) / 2757))^2 / ((566 * 1557) / 2757)
r1c3 <- (468 - ((945 * 1557) / 2757))^2 / ((945 * 1557) / 2757)

r2c1 <- (484 - ((1246 * 1200) / 2757))^2 / ((1246 * 1200) / 2757)
r2c2 <- (239 - ((566 * 1200) / 2757))^2 / ((566 * 1200) / 2757)
r2c3 <- (477 - ((945 * 1200) / 2757))^2 / ((945 * 1200) / 2757)

test.chi <- sum(r1c1, r1c2, r1c3, r2c1, r2c2, r2c3)

1 - pchisq(test.chi, (3 - 1) * (2 - 1))
```

The expected count size is greater than 5 for all cells so a chi-square test can be run.

$H_0$: Gender and Party Identification are Independent
$H_A$: There is a dependence between Gender and Party Identification

$\chi^2 = 30.07$

There is convincing evidence that Gender and Political Affiliation are **dependent** (p-value = 2.954e-07).

# 3

*When drinking tea, a colleague of Ronald Fisher’s at Rothamsted Experiment Station near London claimed she could distinguish whether milk or tea was added to the cup first. To test her claim, Fisher designed an experiment in which she tasted eight cups of tea. Four cups had milk added first, and the other four had tea added first. She was told there were four cups of each type and she should try to select the four that had milk added first. The cups were presented to her in random order. A potential result of the experiment is presented in the table below.*

<Table Not Pictured>

*Perform a test for a positive association between the true order of pouring and her guess. What type of test will you use?*

The Fisher Test should be used to test whether the true odds between Poured First and Guess Poured First are equal to 1.

```{r 3}
tea <- data.frame(Milk=c(3,1), Tea=c(1,3), row.names = c("Milk","Tea"))
fisher.test(tea)
```

There is no convincing evidence that the drink Poured First changed the odds for the First Guess (p-value = 0.4857).
