---
title: 'Homework #1'
author: "Dustin Leatherman"
date: "9/13/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(grid)
library(gridExtra)
library(broom)
library(ggfortify)

ozone <- read.csv("~/Downloads/ozone.csv")
```

# 1

> Create a table with the overall (across sites and days) mean, standard deviation, and percent missing.

```{r}
ozone_pivot <-
  ozone %>%
  # Pivot Day columns to Rows
  pivot_longer(starts_with("Day"), names_to = "day", values_to = "value") %>% 
  mutate(
    # convert day to integer for easier processing
    day = as.integer(str_replace(day, "Day.", ""))
  )

ozone_pivot %>%
  summarise(
    mean = mean(value, na.rm = TRUE), 
    sd = sd(value, na.rm = TRUE), 
    perc_na = sum(is.na(value)) / n()
  ) %>% kable(
    caption = "Overall Summary Statistics"
  ) %>% kable_styling(bootstrap_options = "striped", latex_options = "hold_position")
```

# 2

> a. Compute the mean, variance, and percent missing for each of the n sites; 
> b. Make a histogram of each variable (all three histograms should have n observations); 
> c. create scatter plots of each pair of these variables (each of the three plots should have n points)

```{r 2a}
ozone_summary <-
  ozone_pivot %>% 
  group_by(Station.ID) %>% 
  summarise(
    mean = mean(value, na.rm = TRUE), 
    sd = sd(value, na.rm = TRUE), 
    perc_na = sum(is.na(value)) / n(),
    var = var(value, na.rm = TRUE)
  ) 

ozone_summary %>%
  select(-var) %>% 
  # There are 1000+ stations. Limit for viewing purposes
  head(10) %>% 
  kable(
    caption = "A sample of Station Observation Summaries"
  ) %>% kable_styling(bootstrap_options = "striped", latex_options = "hold_position")
```


```{r 2b, fig.height=3}
plot_mean <- 
  ozone_summary %>% 
  ggplot(aes(x = mean)) + 
    geom_histogram() +
    labs(x = "Mean", y = "Count", title = "Mean")

plot_sd <-
  ozone_summary %>% 
  ggplot(aes(x = sd)) + 
    geom_histogram() +
    labs(x = "Standard Deviation", y = "Count", title = "Std. Deviations")

plot_percmis <-
  ozone_summary %>% 
  ggplot(aes(x = perc_na)) + 
    geom_histogram() +
    labs(x = "Percent Missing", y = "Count", title = "Percent Missing")

grid.arrange(plot_mean, plot_sd, plot_percmis, ncol = 3,  top = textGrob("Ozone Measurements by Station",
                            gp=gpar(fontsize=14,font=1),just=c("center")))
```


```{r 2c, fig.height=3}
plot_mean <- 
  ozone_summary %>% 
  ggplot(aes(x = mean, y = sd)) + 
    geom_point() +
    labs(x = "Mean", y = "Standard Deviation", title = "Std. Deviation vs. Mean")

plot_sd <-
  ozone_summary %>% 
  ggplot(aes(x = sd, y = perc_na)) + 
    geom_point() +
    labs(x = "Standard Deviation", y = "Percent Missing", title = "Missing (%) vs Std. Deviation")

plot_percmis <-
  ozone_summary %>% 
  ggplot(aes(x = mean, y = perc_na)) + 
    geom_point() +
    labs(y = "Percent Missing", x = "Mean", title = "Missing (%) vs. Mean")

grid.arrange(
  plot_mean, 
  plot_sd, 
  plot_percmis, 
  ncol = 3,
  top = textGrob("Ozone Measurement Summaries by Station",
                            gp=gpar(fontsize=14,font=1),just=c("center"))
)
```

# 3

> Conduct a linear regression with response equal to the site's mean and the site's variance and percent missing as covariates.

```{r 3}
model1 <- lm(mean ~ var + perc_na, data = ozone_summary)

tidy(model1) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "hold_position")

autoplot(model1)
```

