---
title: 'Non-linear regression for the motorcycle data'
author: "Phillip Yates - MAT 350 & MAT 427"
date: "10/29/2020"
output:
  pdf_document: default
  html_document: default
---

**This is a useful reference in its entirety so storing it for posterity.**


First we will load the `tidyverse` library for data reshaping and visualization.
```{r,message=FALSE}
library(tidyverse)
```

In this example $X$ is the time since the motorcycle crash and $Y$ is the acceleration of the driver's head. We will fit the semiparametric model
$$Y_{i}\sim\mbox{Normal}\left[g\left(X_{i}\right),\sigma^{2}\right]$$ where the mean function $g$ is assumed to have spline basis representation
$$g(X)=\mu+\sum^{J}_{j=1}B_{j}(X)\beta_{j}.$$ The remaining parameters have uninformative priors: $\mu\sim\mbox{Normal}(0,100)$, $\beta_{j}\sim\mbox{Normal}\left(0,\tau^{2}\right)$ and $\sigma^{2},\tau^{2}\sim\mbox{InvGamma}(0.1,0.1)$.

# Load and plot the motorcycle data

```{r,message=FALSE}
library(MASS)

Y <- mcycle$accel
X <- mcycle$times

Y <- (Y-mean(Y))/sd(Y)
X <- X/max(X)

n <- length(Y)
n

ggplot(NULL,aes(x=X,y=Y))+geom_point()+labs(x="Time",y="Acceleration")
```

# Set up a spline basis expansion

```{r}
library(splines)

# Number of basis functions
J <- 10

# Specify the basis functions -- default is a cubic spline
B <- bs(X,J)
dim(B)
```

```{r}
new.B <- cbind(X,B)
B.data <- as.data.frame(new.B) %>%
  pivot_longer(-X,names_to="Spline",values_to="basis") %>%
  mutate(Spline=factor(Spline,levels=c("1","2","3","4","5","6","7","8","9","10")))
ggplot(B.data,aes(x=X,y=basis,color=Spline))+geom_line()+
  labs(x="Time",y=expression(B[j](X)))
```

# Fit the model

```{r,message=FALSE}
library(rjags)

Moto_model <- "model{
  # Likelihood
    for(i in 1:n){
      Y[i] ~ dnorm(mean[i],taue)
      mean[i] <- mu+inprod(B[i,],beta[])
    }
    
  # Prior
    mu ~ dnorm(0,0.01)
    taue ~ dgamma(0.1,0.1)
    for(j in 1:J){
      beta[j] ~ dnorm(0,taub)
    }
    taub ~ dgamma(0.1,0.1)
}"

dat <- list(Y=Y,n=n,B=B,J=J)
init <- list(mu=mean(Y),beta=rep(0,J),taue=1/var(Y))
model <- jags.model(textConnection(Moto_model),inits=init,data=dat,quiet=TRUE)
update(model,10000,progress.bar="none")

samp <- coda.samples(model,variable.names=c("mean"),n.iter=20000,progress.bar="none")
```

# Plot the fixed curve, $g(X)$

```{r}
sum <- summary(samp)
names(sum)
```

```{r}
q <- sum$quantiles
q.small <- cbind(X,q[,c(1,3,5)])
q.data <- as.data.frame(q.small) %>%
  pivot_longer(-X,names_to="Method",values_to="estimate") %>%
  mutate(Method=ifelse(Method=="50%","Median",
                ifelse(Method=="2.5%","Lower: 95% Interval","Upper: 95% Interval")))

ggplot(NULL,aes(x=X,y=Y))+geom_point()+
  geom_line(data=q.data,aes(x=X,y=estimate,linetype=Method),color="red")+
  scale_linetype_manual(values=c(2,1,2),
                        labels=c("Lower: 95% Interval","Median","Upper: 95% Interval"))+labs(x="Time",y="Acceleration",linetype="")
```

**Summary:** The mean trend seems to fit the data well; however, the variance of the observations around the mean varies with $X$.

# Heteroskedastic model

The variance is small for $X$ near zero and increases with $X$. To account for this, we allow the log of the variance to vary with $X$ following a second spline basis expansion:
$$Y_{i}\sim\mbox{Normal}\left[g\left(X_{i}\right),\sigma^{2}\left(X_{i}\right)\right]$$ where $g(X)=\mu+\sum^{J}_{i=1}B_{j}(X)\beta_{j}$ is modeled as above and
$$\log\left[\sigma^{2}(X)\right]=\mu_{2}+\sum^{J}_{i=1}B_{j}(X)\alpha_{j}.$$ The parameters have uninformative priors $\mu_{k}\sim\mbox{Normal}(0,100)$, $\beta_{j}\sim\mbox{Normal}\left(0,\sigma_{b}^{2}\right)$, $\alpha_{j}\sim\mbox{Normal}\left(0,\sigma_{a}^{2}\right)$, and $\sigma_{a}^{2},\sigma_{b}^{2}\sim\mbox{InvGamma}(0.1,0.1)$.

```{r}
moto_model2 <- "model{
  # Likelihood
    for(i in 1:n){
      Y[i] ~ dnorm(mean[i],inv_var[i])
      mean[i] <- mu1+inprod(B[i,],beta[])
      inv_var[i] <- 1/sig2[i]
      log(sig2[i]) <- mu2+inprod(B[i,],alpha[])
    }
  
  # Prior
    mu1 ~ dnorm(0,0.01)
    mu2 ~ dnorm(0,0.01)
    for(j in 1:J){
      beta[j] ~ dnorm(0,taub)
      alpha[j] ~ dnorm(0,taua)
    }
    taua ~ dgamma(0.1,0.1)
    taub ~ dgamma(0.1,0.1)
    
  # Prediction intervals
    for(i in 1:n){
      low[i] <- mean[i]-1.96*sqrt(sig2[i])
      high[i] <- mean[i]+1.96*sqrt(sig2[i])
    }
}"
```

# Fit the model

```{r}
dat <- list(Y=Y,n=n,B=B,J=J)
init <- list(mu1=mean(Y),beta=rep(0,J),mu2=log(var(Y)),alpha=rep(0,J))
model <- jags.model(textConnection(moto_model2),inits=init,data=dat,quiet=TRUE)

update(model,10000,progress.bar="none")

samp2 <- coda.samples(model,variable.names=c("mean","sig2","low","high"),
                      n.iter=20000,progress.bar="none")
```

# Plot the fixed curve, $g(X)$

```{r}
q2 <- summary(samp2)$quantiles
q2.small.mean <- cbind(X,q2[1:n+2*n,c(1,3,5)])

q2.data.mean <- as.data.frame(q2.small.mean) %>%
  pivot_longer(-X,names_to="Method",values_to="estimate") %>%
  mutate(Method=ifelse(Method=="50%","Median",
                ifelse(Method=="2.5%","Lower: 95% Interval","Upper: 95% Interval")))

ggplot(NULL,aes(x=X,y=Y))+geom_point()+
  geom_line(data=q2.data.mean,aes(x=X,y=estimate,linetype=Method),color="red")+
  scale_linetype_manual(values=c(2,1,2),
                        labels=c("Lower: 95% Interval","Median","Upper: 95% Interval"))+labs(x="Time",y="Acceleration",linetype="")
```

We can also plot the fitted variance function.
```{r}
q2.small.sig <- cbind(X,q2[1:n+3*n,c(1,3,5)])

q2.data.sig <- as.data.frame(q2.small.sig) %>%
  pivot_longer(-X,names_to="Method",values_to="estimate") %>%
  mutate(Method=ifelse(Method=="50%","Median",
                ifelse(Method=="2.5%","Lower: 95% Interval","Upper: 95% Interval")))

ggplot(q2.data.sig,aes(x=X,y=estimate,linetype=Method))+geom_line(color="red")+
  scale_linetype_manual(values=c(2,1,2),
                        labels=c("Lower: 95% Interval","Median","Upper: 95% Interval"))+labs(x="Time",y="Variance",linetype="")
```

Finally we can plot the prediction intervals.
```{r}
q2.small.low <- cbind(X,q2[1:n+1*n,3])
q2.small.high <- cbind(X,q2[1:n+0*n,3])

ggplot(NULL,aes(x=X,y=Y))+geom_point()+
  geom_line(data=as.data.frame(q2.small.low),aes(x=X,y=V2),color="red")+
  geom_line(data=as.data.frame(q2.small.high),aes(x=X,y=V2),color="red")+
  labs(x="Time",y="Acceleration")
```

