---
title: 'Quiz #3'
author: "Dustin Leatherman"
date: "5/17/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(tidyverse)
library(broom)
```

# 1 

*For the 23 space shuttle flights before the Challenger mission disaster in 1986, the data set Oring.csv shows the temperature (◦F) at the time of the flight and whether at least one primary O-ring suffered thermal distress.*

```{r intro}
oring <- read.csv("~/Downloads/Oring.csv")
head(oring) %>% tidy %>% kable
```

## a

*Use logistic regression to model the effect of temperature on the probability of thermal distress. Interpret the effect.*

```{r 1a}
oring.lrm <- glm(TD ~ Tempeature, family = "binomial", data = oring)
tidy(oring.lrm)
```

There is moderate evidence that temperature is associated with thermal distress (Two-Tailed Wald Logistic Regression on a Single Variable. p-value = 0.032).

## b 

*Estimate the probability of thermal distress at 31◦F , the temperature at the time of the Challenger flight.*

The estimated logit of failure probablitiy is 1.7164. The estimated failure probability is 0.9962.

```{r 1b}
eta <- predict(oring.lrm, newdata = data.frame(Tempeature=c(31)))
est.logit <- log(eta)
est.prob <- exp(eta) / (1 + exp(eta))

est.logit
est.prob
```

## c

*At what temperature does the estimated probability equal 0.50? At that temperature, give a linear approximation for the change in the estimated probability per degree increase in temperature. Use $\hat{\beta_1}\pi(1 - \pi)$ for the second part of this problem.*

$\hat{y} = 15.0429 - 0.2322 \times Temperature$

$logit(0.50) = 15.0429 - 0.2322 \times Temperature$

$\to logit(0.50 = log(0.5/0.5) = log(1) = 0$

$\to 0 = 15.0429 - 0.2322 \times Temperature \to Temperature = \frac{15.0429}{0.2322} = 64.7842$ 

### Linear Approximation

$0.5 \times 0.5 \times -0.2322 = -0.0581$

## d

*Interpret the effect of temperature on the odds of thermal distress.*

For every 1 degree increase in temperature, the odds of thermal distress vs non-thermal distress decrease by 0.0581.

## e

*Test the hypothesis that temperature has no effect using the Wald test and using the drop-in-deviance test.*

There is convincing evidence that temperature has no effect on O-ring failure (Drop-in-deviance Test. p-value =  0.167).

There is moderate evidence that temperature has an effect on O-ring failure (Two-Tailed Wald Logistic Regression on a Single Variable. p-value = 0.032).

Given that the drop-in-deviance test more powerful than the Wald Test, the conclusions from the Drop-in-Deviance test are preferred.

```{r 1e}
# Drop in Deviance Test
1 - pchisq(oring.lrm$null.deviance, oring.lrm$df.null)
```
