---
title: 'Homework #1'
author: "Dustin Leatherman"
date: "1/11/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  fig.width=8, fig.height=3)

library(ggplot2)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)

pdiff <- function(ls) {
  a <- ls[1]
  b <- ls[2]
  (abs(a - b) / ((a + b) / 2)) * 100
}
```

# 1.5
> For a moving average process of the form
> $x_t = w_{t - 1} + 2 w_t + w_{t + 1}$ 

> where $w_t$ are independent with zero means and variance $\sigma_w^2$ , determine the autocovariance and autocorrelation functions as a function of lag h and sketch the ACF as a function of h.

$$
\begin{aligned}
\hat{\gamma(h)} &= cov(x_t, x_{t + h}) = E(x_t \times x_{t + h})\\
&= E[(w_{t - 1} + 2 w_t + w_{t + 1})(w_{t + h - 1} + 2 w_{t + h} + w_{t + h + 1})]\\
&= E[w_{t - 1} w_{t + h - 1} + 2 w_{t - 1} w_{t + h} + w_{t - 1} w_{t + h + 1} + 2 w_{t} w_{t + h - 1} + 4 w_{t} w_{t + h} + 2 w_{t} w_{t + h + 1} + w_{t + 1} w_{t + h - 1} + 2 w_{t + 1} w_{t + h} + w_{t + 1} w_{t + h + 1}]
\end{aligned}
$$
$\hat{\rho(h)} = \frac{\hat{\gamma(h)}}{\hat{\gamma(0)}} = \frac{\hat{\gamma}(h)}{\sigma_w^2}$

Note the following definitions
$$
\begin{aligned}
E(w_t^2) &= var(w_t) + E(w_t)^2\\
E(w_t) &= 0\\
E(w_t w_s) &= 0, \ \text{where} \ s \neq t\\
var(w_t) &= \sigma_w^2\\
E(w_t^2) &= \sigma_w^2
\end{aligned}
$$

In the interest of brevity and readability, the following Expectations only contain non-zero terms. 

When |h| = 0:

$$
E[x_t^2] = E[w_{t - 1}^2 + 4 w_t^2 + w_{t + 1}^2] = \sigma_w^2 + 4 \sigma_w^2 + \sigma_w^2 = 6 \sigma_w^2
$$

When |h| = 1:

$$
E[x_t x_{t + 1}] = E[2 w_t^2 + 2 w_{t + 1}^2] = 4 \sigma_w^2
$$

When |h| = 2:

$$
E[x_t x_{t + 2}] = E(w_{t + 1}^2) = \sigma_w^2
$$

When |h| = 3:

$$
E[x_t x_{t + 3}] = 0
$$

```{r}
data <- data.frame(h = c(0,1,2,3), acf = c(6,4,1,0))
ggplot(data, aes(x = h, y = acf)) + 
  geom_bar(stat = "identity", fill = "blue")
```

# 1.12

> Let $w_t$ , for $t = 0, \pm 1, \pm 2$,...$ be a normal white noise process, and consider
> the series
> $x_t = w_t w_{t - 1}$
> Determine the mean and autocovariance function of $x_t$, and state whether it is stationary.

Given $w_t \sim ind \ N(0, \sigma_w^2)$.
$$
\mu_{x_t} = E(x_t) = E(w_t w_{t - 1}) = E(w_i) * E(w_{t - 1}) = 0
$$

$$
\begin{aligned}
\hat{\gamma}(h) &= E(x_t \times x_{t + h})\\
&= E(w_t w_{t - 1} w_{t + h} w_{t + h - 1})\\
&= E(w_t w_{t - 1}) \times E(w_{t + h} w_{t + h - 1})\\
&= 0
\end{aligned}
$$

Since this time series is the product of two stationary time series, it is considered stationary. This is further proven by the mean and covariance functions being independent of time.

# 1.14

> (a) Simulate a series of n = 500 Gaussian white noise observations as in Example 1.6 
> and compute the sample ACF, $\hat{\rho}(h)$ to lag 20. 
> Compare the sample ACF you obtain to the actual ACF, $\rho(h)$. [Recall Example 1.18.]
> (b) Repeat part (a) using only n = 50. How does changing n affect the results?

## a

```{r}
set.seed(123)

w500 <- rnorm(500)

w500.acf <- 
  bind_cols(
  acf(w500, type="correlation", plot = FALSE) %>% 
    tidy %>% 
    select(lag, correlation = acf),
  acf(w500, type="covariance", plot = FALSE) %>% 
    tidy %>% 
    select(covariance = acf)
) 

w500.acf %>% 
  filter(lag <= 20) %>%
  kable(
    caption = "First 20 lags of 500 random Normal observations"
  ) %>% 
  kable_styling(full_width = FALSE, protect_latex = TRUE, latex_options = "hold_position")

ggplot(w500.acf, aes(x = lag)) + 
  geom_line(aes(y = correlation), linetype = "dotted") + 
  geom_line(aes(y = covariance, color = "red"), linetype = "dotted", show.legend = FALSE)
```



The actual correlation coefficient for Normal White Noise is equal to the covariance between a given time point and lag $h$. The values are very close as seen in the table and the graph.

## b

```{r}
w50 <- rnorm(50)
w50.acf <-
  bind_cols(
  acf(w50, type="correlation", plot = FALSE) %>% 
    tidy %>% 
    select(lag, correlation = acf),
  acf(w50, type="covariance", plot = FALSE) %>% 
    tidy %>% 
    select(covariance = acf)
) 

w50.acf %>%
  filter(lag <= 20) %>% 
  kable(
    caption = "Lags of 50 random Normal observations"
  ) %>% 
  kable_styling(full_width = FALSE, protect_latex = TRUE, latex_options = "hold_position")

ggplot(w50.acf, aes(x = lag)) + 
  geom_line(aes(y = correlation), linetype = "dotted") + 
  geom_line(aes(y = covariance, color = "red"), linetype = "dotted", show.legend = FALSE)
```

Changing n reduces the number of lags that are created and the sample correlation drifts further from the actual correlation. This is an expected effect when using smaller sample sizes.